A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-01-08 doc/large_strain.tex
remove constants from operation counts

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the residual force vector $\gz F$ corresponding to the discretization are assembled (that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored) and a direct or iterative solver is used to solve the linear system.
In this case the the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
On modern computer architectures however, it transpires that loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
The idea is to perform the operations within the solver on-the-fly rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
%
The matrix-vector product $\gz A \gz x$ (i.e. the action of operator $\gz A$ on a vector $\gz x$) can be expressed by
\begin{align}
  \begin{split}
 (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &\approx \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K_q
  \end{split}
  \label{eq:mf_vmult}
\end{align}
where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ and
$w_q$ is the quadrature weight.
For the sake of demonstration, let us assume that $\gz A$ represents a mass operator with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
\begin{align*}
 \tilde{a}(N_i,N_j)(\gz \xi_q) = N_i(\gz \xi_q)N_j(\gz \xi_q).
\end{align*}
Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in $d$ space dimensions.
For the total matrix-vector product we get $\mathcal{O}(2n^{2d})$,
and this is the same as for a regular matrix-vector product where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.

@@

Let us illustrate this in 2D. Tensor-product quadrature points can be expressed as a combination of one dimensional quadrature points
\begin{align}
  \gz \xi_q = (\widetilde{\xi}_{q_1}, \widetilde{\xi}_{q_2}),
\end{align}
where for each quadrature points $q$ we can associate a multi-index $(q_1,q_2)$.
Weights associated with the quadrature formula $w_q$ can be expressed as a product of one-dimensional weights
\begin{align}
  w_q = \widetilde{w}_{q_1} \widetilde{w}_{q_2} \, .
\end{align}
Shape functions can also be expressed as product of one dimensional basis functions
\begin{align}
  N_i(\gz \xi_q) =
  \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1})
  \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2})
\end{align}
where for each DoF index $i$ we can associate a multi-index $(i_1, i_2)$.
See Figure \ref{fig:multi_index} for an illustration.
%
Therefore, the result of the application of the scalar-valued mass operator, represented as a vector $y_i$, on a single element $K$ can be written as
\begin{align*}
  y_{i} \equiv Y_{i_1\,i_2} & = \sum_{q_1} \sum_{q_2} \sum_{j_1} \sum_{j_2}
  \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1})
  \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2})
  \widetilde{N}_{j_1}(\widetilde{\xi}_{q_1})
  \widetilde{N}_{j_2}(\widetilde{\xi}_{q_2})
  X_{j_1\,j_2}
  \widetilde{w}_{q_1} \widetilde{w}_{q_2} J^K_{q_1\,q_2}
  \\
  &=
  \sum_{q_1} \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1}) \widetilde{w}_{q_1}
  \sum_{q_2} \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2}) \widetilde{w}_{q_2}
  J^K_{q_1\,q_2}
  \left[
    \sum_{j_1}
    \widetilde{N}_{j_1}(\widetilde{\xi}_{q_1})
    \sum_{j_2}
    \widetilde{N}_{j_2}(\widetilde{\xi}_{q_2})
    X_{j_1\,j_2}
  \right]\, \forall i_1 \, i_2
\end{align*}
%
Note that $x_j \equiv X_{j_1\,j_2}$; the former accesses the element source vector using a linear index $j$, whereas the latter uses the associated multi-index accesses $\{j_1\,j_2\}$.
%
These four loops all have the same structure.
We either keep the shape function fixed and iterate over the quadrature points in one spatial direction (in the first two sums) or keep the quadrature point fixed and iterate over all the shape functions in one spatial direction (in the last two sums).
Since each of these $2\times d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, this results in an algorithm with an arithmetic complexity of $\mathcal{O}(2dn^{d+1})$.
Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for the 3D case, and where the discretization and evaluation employs high polynomial degree basis functions with the corresponding quadrature rule.
More information on matrix-free techniques can be found in \cite{kronbichler12,vos10}.

>>>>

\section{Matrix-free operator evaluation}
\label{sec:mf}
Classically, in order to solve \eqref{eq:linear_system} the matrix $\gz A$ and the residual force vector $\gz F$ corresponding to the discretization are assembled (that is, the non-zero matrix elements $A_{ij}$ are individually computed and stored) and a direct or iterative solver is used to solve the linear system.
In this case the the matrix-vector product $\gz A \gz x$ results from the product of the individual matrix elements with the elements in the vector, and is usually implemented with the aid of specialized linear algebra packages.
On modern computer architectures however, it transpires that loading sparse matrix data into the CPU registers is significantly slower than performing the arithmetic operations when solving.
For this reason, recent implementations often focus on so-called matrix-free approaches where the matrix is not assembled but rather the result of its operation on a vector is directly calculated.
The idea is to perform the operations within the solver on-the-fly rather than loading matrix elements from memory.
For iterative solvers, this is possible since it is sufficient to compute matrix-vector products alone.
%
The matrix-vector product $\gz A \gz x$ (i.e. the action of operator $\gz A$ on a vector $\gz x$) can be expressed by
\begin{align}
  \begin{split}
 (\gz A \gz x)_i &= \sum_j a(\gz N_i,\gz N_j) x_j \\
        &\approx \sum_K \sum_q \sum_j \tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q) x_j w_q J^K_q
  \end{split}
  \label{eq:mf_vmult}
\end{align}
where $\tilde{a}(\gz N_i,\gz N_j)(\gz \xi_q)$ is a representation for the evaluation of the bilinear form in one quadrature point $\gz \xi_q$, $J^K_q$ is the Jacobian of the mapping from the isoparametric element to the element $K$ and
$w_q$ is the quadrature weight.
For the sake of demonstration, let us assume that $\gz A$ represents a mass operator with scalar valued shape functions $\{ N_i(\gz x) \}$. In this case, it holds that
\begin{align*}
 \tilde{a}(N_i,N_j)(\gz \xi_q) = N_i(\gz \xi_q)N_j(\gz \xi_q).
\end{align*}
Further assuming that the number of quadrature points $n$ in one-dimension matches the degree of the ansatz space plus one, the evaluation of all of the shape functions ($n^d$)
in all quadrature points ($n^d$) has complexity $\mathcal{O}(n^{2d})$ in $d$ space dimensions.
For the total matrix-vector product we get $\mathcal{O}(n^{2d})$,
and this is the same as for a regular matrix-vector product where the entries are already precomputed. Assuming tensor product ansatz spaces and a tensor product quadrature rule, evaluation of this operations can be performed more efficiently using a technique called ``sum factorization''.

@@

Let us illustrate this in 2D. Tensor-product quadrature points can be expressed as a combination of one dimensional quadrature points
\begin{align}
  \gz \xi_q = (\widetilde{\xi}_{q_1}, \widetilde{\xi}_{q_2}),
\end{align}
where for each quadrature points $q$ we can associate a multi-index $(q_1,q_2)$.
Weights associated with the quadrature formula $w_q$ can be expressed as a product of one-dimensional weights
\begin{align}
  w_q = \widetilde{w}_{q_1} \widetilde{w}_{q_2} \, .
\end{align}
Shape functions can also be expressed as product of one dimensional basis functions
\begin{align}
  N_i(\gz \xi_q) =
  \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1})
  \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2})
\end{align}
where for each DoF index $i$ we can associate a multi-index $(i_1, i_2)$.
See Figure \ref{fig:multi_index} for an illustration.
%
Therefore, the result of the application of the scalar-valued mass operator, represented as a vector $y_i$, on a single element $K$ can be written as
\begin{align*}
  y_{i} \equiv Y_{i_1\,i_2} & = \sum_{q_1} \sum_{q_2} \sum_{j_1} \sum_{j_2}
  \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1})
  \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2})
  \widetilde{N}_{j_1}(\widetilde{\xi}_{q_1})
  \widetilde{N}_{j_2}(\widetilde{\xi}_{q_2})
  X_{j_1\,j_2}
  \widetilde{w}_{q_1} \widetilde{w}_{q_2} J^K_{q_1\,q_2}
  \\
  &=
  \sum_{q_1} \widetilde{N}_{i_1}(\widetilde{\xi}_{q_1}) \widetilde{w}_{q_1}
  \sum_{q_2} \widetilde{N}_{i_2}(\widetilde{\xi}_{q_2}) \widetilde{w}_{q_2}
  J^K_{q_1\,q_2}
  \left[
    \sum_{j_1}
    \widetilde{N}_{j_1}(\widetilde{\xi}_{q_1})
    \sum_{j_2}
    \widetilde{N}_{j_2}(\widetilde{\xi}_{q_2})
    X_{j_1\,j_2}
  \right]\, \forall i_1 \, i_2
\end{align*}
%
Note that $x_j \equiv X_{j_1\,j_2}$; the former accesses the element source vector using a linear index $j$, whereas the latter uses the associated multi-index accesses $\{j_1\,j_2\}$.
%
These four loops all have the same structure.
We either keep the shape function fixed and iterate over the quadrature points in one spatial direction (in the first two sums) or keep the quadrature point fixed and iterate over all the shape functions in one spatial direction (in the last two sums).
Since each of these $2\times d$ loops has a complexity of $n$ and we perform them for $n^d$ values in quadrature points or values in degrees of freedoms simultaneously, this results in an algorithm with an arithmetic complexity of $\mathcal{O}(n^{d+1})$.
Hence, we can expect that such a matrix-free approach is faster than a regular matrix-based approach, especially for the 3D case, and where the discretization and evaluation employs high polynomial degree basis functions with the corresponding quadrature rule.
More information on matrix-free techniques can be found in \cite{kronbichler12,vos10}.
